---
title: "Research Methods - Assignment 1"
author: "Dean Baltiansky"
date: "10/28/2021"
output: pdf_document
---

```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(apaTables)
library(knitr)
library(kableExtra)
library(tinytex)
library(papaja)
library(formatR)
```

```{r, echo = F}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

# 1 Load data

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
df_exp1 <- read.csv("https://raw.githubusercontent.com/bocowgill-collaborations/ResearchMethods-Repository/master/HW1/assignment1-research-methods.csv",sep = "")
```

# 2 Statistical analyses

Does having an education from an elite college increase the chances of getting
a callback?

```{r}
m1 <- lm(calledback ~ EliteSchoolCandidate, data = df_exp1)
```

# 3 Output

```{r, results = "asis"}
apa_lm <- apa_print(m1)
apa_table(
  apa_lm$table, 
  caption = "Linear model predicting chance of receiving a call back for an interview from having a degree at an elite university",
  placement = "h"
)
```

The analysis indicates a positive relationship between having a degree at an elite university and getting a call back for an interview. That is, when sending out resumes that are otherwise identical, having a degree at an elite university increases the chances of getting a call back for an interview.
